Anomaly detection with kernel density estimation on manifolds

### 2022-02-26 paper/paper.Rmd
Application data introduction and steps

<!-- Dataset -->

<!-- Goal and process -->

<!-- Results (Plots and tables) -->

<!-- Analysis and summary -->

@@

[TODO]

Next, we consider smart-meter data for residential and non-profiled meter consumers, collected in the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data]. Electricity smart meters record consumption, on a near real-time basis, at the level of individual commercial and residential properties. The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] does not include energy for heating systems since it is either metered separately, or households use a different source of energy, such as oil or gas. In this study, the installed cooling systems are also not reported.

@@

We use measurements of half-hourly electricity consumption gathered from 3,639 residential consumers over 535 consecutive days. Every meter provides electricity consumption between 14 July 2009 and 31 December 2010. Demand data from two smart meters (ID 1003 and 1539) are shown in \autoref{fig:smartmeter} as time series plots. It is obvious that these meters have relatively different patterns. Meter 1539 (bottom of \autoref{fig:smartmeter}) has a period of around 150 days with lower (approximately half) the electricity usage of remaining days and is otherwise relatively stable. In contrast, Meter 1003 (top of \autoref{fig:smartmeter}) exhibits regular spikes on weekends.

For electricity demand data, one particular problem of interest is to visualize and identify households or periods of the week with anomalous usage patterns. For this reason, the object of interest in comparing households or periods of the week is the distribution of electricity demand rather than the raw data itself [@Hyndman2018-ia]. An additional advantage of looking at distributions rather than raw data is that it provides a convenient mechanism for handling missing data which can be a significant problem in smart meter data applications. 
In the rest of this section, we first describe how discrete approximations to the distributions of interest are computed. We then approximate the distances between vectors containing probabilities with the consistent estimator in \autoref{estimator}, which allows us to apply manifold learning techniques on statistical manifolds.

<!-- ## Data processing {#dataprocessing} -->

Let $d_{i,t}$ denote the electricity demand for observation $i$ and for time period $t$ (subsequently we will see that $i$ can either index the household, time of the week, or both, while $t$ may index the week or half-hour period). The objective is to estimate the distances between the distribution of electricity demand for observation $i$, $F_i$, over time. Since the raw data could be thought of as samples generated from the true electricity usage distributions for different households and different periods, we could apply the consistent Hellinger distance estimator derived in \autoref{estimator} to the smart meter data.
The first step is to group all data together regardless of $i$ and $t$, and take the data as samples from the reference probability distribution $R$. The sample size $r=93,616,914$. Next, we construct $T_r=100$ intervals with about $1\%$ samples from $R$ within each interval (except for the last interval). The segments $\{I_g^r\}_{g=1,2,\dots,100}$ can then be used to construct discrete distribution approximation to $F_i$ by counting the ratio of samples points from $F_i$ fallen into segment $I_g^r$.
This is found by computing $\pi_{i,g}=(1/m)\sum_t I(d_{i,t} \in I_g^r)$ where $m$ is the sample size from $F_i$. 
Then the Hellinger distance between $F_i$ and $F_j$ could be estimated using Equation \@ref(eq:hellingerest1).
In this way, we make it possible to search nearest neighbors more efficiently using ANN methods and further apply manifold learning algorithms.

@@

Summary of the contribution and conclusion

Details about the process

Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).

Discuss simulation with non-uniform meta data.

>>>>

@@

In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010. The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems. We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.

For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia]. @Cheng2021-ex propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution. @Cheng2021-ex use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher. By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.

 In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households. Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers. The data processing steps has been clearly stated in the application section of @Cheng2021-ex and they are skipped here. Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}. However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.

@@

@@

<!-- Summary of the contribution and conclusion -->

<!-- Details about the process -->

<!-- Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate). -->

<!-- Discuss simulation with non-uniform meta data. -->
